---
title: "Simulation-Based Verification for NFL WR 40-yard Dash Extremes"
author: "Francis Yap"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo    = TRUE,
message = TRUE,
warning = TRUE
)
```
Overview

This document provides a reproducible simulation-based verification of generalized extreme-value (GEV) modeling for NFL wide receiver (WR) 40-yard dash performance.

The analysis proceeds as follows:

1. NFL Combine WR 40-yard dash times are read from Sports-Reference HTML tables saved with an .xls extension (one file per year).

2. Annual block minima of dash times are computed and analyzed on the negative-time scale (block maxima of -time).

3. GEV and Gumbel models are optionally fit to empirical annual maxima as a diagnostic check.

4. A Monte Carlo study evaluates the finite-sample behavior of GEV return-level estimation, including bias, RMSE, and empirical convergence rates.

5. Return levels are transformed to the speed scale post-estimation, and Delta-method implications are examined.

Packages
```{r}
suppressPackageStartupMessages({
library(xml2)
library(rvest)
library(dplyr)
library(purrr)
library(extRemes)
library(e1071)
library(ggplot2)
library(tidyr)
library(readr)
library(scales)
})
```
Data input and helper functions
```{r}
folder_path <- "C:/Users/yapfr/Downloads/NFL Project"

file_list <- list.files(
  path       = folder_path,
  pattern    = "^sportsref_download_\\d{4}\\.xls$",
  full.names = TRUE
)

if (length(file_list) == 0) {
  stop("No files found. Check folder_path and filename pattern.")
}

# Read Sports-Reference HTML disguised as .xls and return the first HTML table

read_sportsref_xls_html <- function(fpath) {
  doc  <- read_html(fpath)
  tabs <- html_table(doc, fill = TRUE)
  
  if (length(tabs) == 0) stop("No tables found in file: ", fpath)
  
  tabs[[1]]
}

# Extract year from filename sportsref_download_YYYY.xls

extract_year_from_filename <- function(fpath) {
  as.numeric(gsub(".*_(\\d{4})\\.xls$", "\\1", basename(fpath)))
}

# Robust numeric conversion (handles "", "--", etc.)

to_numeric_quiet <- function(x) suppressWarnings(as.numeric(x))
```
Load WR 40-yard dash times
```{r}
all_wr_times <- map_dfr(file_list, function(fpath) {
yr <- extract_year_from_filename(fpath)
df <- read_sportsref_xls_html(fpath)

df %>%
filter(.data$Pos == "WR") %>%
mutate(time_40 = to_numeric_quiet(.data$`40yd`)) %>%
filter(!is.na(.data$time_40)) %>%
transmute(year = yr, time_40)
})

print(head(all_wr_times, 20))
cat("\nTotal rows:", nrow(all_wr_times), "\n")
summary(all_wr_times$time_40)
```
Parent distribution calibration
```{r}
mu_T  <- mean(all_wr_times$time_40, na.rm = TRUE)
sd_T  <- sd(all_wr_times$time_40,   na.rm = TRUE)
var_T <- var(all_wr_times$time_40,  na.rm = TRUE)

counts <- as.numeric(table(all_wr_times$year))

skew_T <- e1071::skewness(all_wr_times$time_40, na.rm = TRUE)
k      <- 8 / (skew_T^2)

cat("\nEmpirical parent moments (time scale):\n")
print(c(mu_T = mu_T, sd_T = sd_T, var_T = var_T))

cat("\nPer-year WR counts summary:\n")
print(summary(counts))
cat("Median:", median(counts), " SD:", sd(counts), " Var:", var(counts), "\n")

cat("\nSkewness(time_40):", skew_T, "\n")
cat("k = 8 / skew^2:", k, "\n")
```
Fixed calibration for reproducibility

For the Monte Carlo study, the parent distribution is fixed to the calibration reported in the manuscript to ensure reproducibility.
```{r}
freeze_parent <- TRUE

if (isTRUE(freeze_parent)) {
mu_T       <- 4.5072520
sd_T       <- 0.1017394
n_per_year <- 44L
} else {
n_per_year <- as.integer(median(counts))
}

cat("\nParent parameters used in Monte Carlo:\n")
print(c(mu_T = mu_T, sd_T = sd_T, n_per_year = n_per_year))
```
Empirical annual maxima

Annual block minima of time correspond to block maxima of -time. We also track maxima on the speed scale.
```{r}
annual_maxima <- map_dfr(file_list, function(fpath) {
yr <- extract_year_from_filename(fpath)
df <- read_sportsref_xls_html(fpath)

wr <- df %>%
filter(.data$Pos == "WR") %>%
mutate(time_40 = to_numeric_quiet(.data$`40yd`)) %>%
filter(!is.na(.data$time_40))

if (nrow(wr) == 0) {
return(tibble(year = yr, max_neg_time = NA_real_, max_speed = NA_real_))
}

tibble(
year         = yr,
max_neg_time = max(-wr$time_40,      na.rm = TRUE),
max_speed    = max(40 / wr$time_40,  na.rm = TRUE)
)
}) %>%
arrange(year)

print(head(annual_maxima, 20))
```
Optional GEV vs Gumbel diagnostic fits
```{r}
gev_neg <- fevd(x = annual_maxima$max_neg_time, type = "GEV",    method = "MLE")
gum_neg <- fevd(x = annual_maxima$max_neg_time, type = "Gumbel", method = "MLE")

gev_speed <- fevd(x = annual_maxima$max_speed, type = "GEV",    method = "MLE")
gum_speed <- fevd(x = annual_maxima$max_speed, type = "Gumbel", method = "MLE")

cat("\n--- Empirical fits: -time scale ---\n")
print(summary(gev_neg))
print(summary(gum_neg))

cat("\n--- Empirical fits: speed scale ---\n")
print(summary(gev_speed))
print(summary(gum_speed))
```
Diagnostic plots
```{r}
par(mar = c(4, 4, 2, 1))
plot(gev_neg)
plot(gum_neg)
plot(gev_speed)
plot(gum_speed)
```

Likelihood-ratio tests (xi = 0 vs free xi)
```{r}
ll_gev_neg <- -gev_neg$results$value
ll_gum_neg <- -gum_neg$results$value
LRT_neg    <- 2 * (ll_gev_neg - ll_gum_neg)
pval_neg   <- 1 - pchisq(LRT_neg, df = 1)

ll_gev_speed <- -gev_speed$results$value
ll_gum_speed <- -gum_speed$results$value
LRT_speed    <- 2 * (ll_gev_speed - ll_gum_speed)
pval_speed   <- 1 - pchisq(LRT_speed, df = 1)

cat("\nLRT (xi=0 vs free):\n")
print(c(LRT_neg = LRT_neg, pval_neg = pval_neg, LRT_speed = LRT_speed, pval_speed = pval_speed))
```
Monte Carlo simulation
```{r}
set.seed(20251122)

years_grid <- c(5, 10, 15, 20, 25, 26, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75,
80, 85, 90, 95, 100)

n_sim   <- 1000
periods <- years_grid
```
Simulation helper functions (parallel to the R script)
```{r}
simulate_block_maxima_normal <- function(n_years, n_per_year, mu_T, sd_T) {
times <- rnorm(n_years * n_per_year, mean = mu_T, sd = sd_T)
mat   <- matrix(times, nrow = n_years, ncol = n_per_year, byrow = TRUE)

year_min_time <- apply(mat, 1, min)   # annual minima of time
neg_time      <- -year_min_time       # annual maxima on -time scale

list(neg_time = neg_time)
}

fit_gev_and_RL <- function(x, periods) {
tryCatch({
fit <- suppressWarnings(fevd(x, type = "GEV"))
rl  <- suppressWarnings(return.level(fit, return.period = periods, do.ci = FALSE))

pars <- fit$results$par
list(
  mu    = unname(pars["location"]),
  sigma = unname(pars["scale"]),
  xi    = unname(pars["shape"]),
  RL    = as.numeric(rl)
)

}, error = function(e) {
list(
mu    = NA_real_,
sigma = NA_real_,
xi    = NA_real_,
RL    = rep(NA_real_, length(periods))
)
})
}

simulate_once <- function(sim_id, n_years, n_per_year, mu_T, sd_T, periods) {
sim <- simulate_block_maxima_normal(
n_years    = n_years,
n_per_year = n_per_year,
mu_T       = mu_T,
sd_T       = sd_T
)

fit_neg <- fit_gev_and_RL(sim$neg_time, periods = periods)

# Transform RLs to speed AFTER estimation

RL_speed <- 40 / (-fit_neg$RL)

tibble(
sim_id    = sim_id,
years     = n_years,
mu_hat    = fit_neg$mu,
sigma_hat = fit_neg$sigma,
xi_hat    = fit_neg$xi,
RL_period = periods,
RL_neg    = fit_neg$RL,
RL_speed  = RL_speed
)
}
```
Monte Carlo execution
```{r}
results_long <- map_dfr(years_grid, function(n_years) {
message("Simulating for n = ", n_years, " years ...")
map_dfr(seq_len(n_sim), function(s) {
simulate_once(
sim_id     = s,
n_years    = n_years,
n_per_year = n_per_year,
mu_T       = mu_T,
sd_T       = sd_T,
periods    = periods
)
})
})

dplyr::glimpse(results_long)

nf_counts <- results_long %>%
summarise(
n_bad_RL_neg   = sum(!is.finite(RL_neg)),
n_bad_RL_speed = sum(!is.finite(RL_speed)),
n_bad_mu       = sum(!is.finite(mu_hat)),
n_bad_sigma    = sum(!is.finite(sigma_hat)),
n_bad_xi       = sum(!is.finite(xi_hat))
)

cat("\nNon-finite counts:\n")
print(nf_counts)
```
Boxplot diagnostics (zoomed)
```{r}
plot_RL_box_zoom <- function(df_long,
RLp,
which_scale = c("neg", "speed"),
zoom_prob = 0.99,
x_breaks = c(5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)) {
which_scale <- match.arg(which_scale)

dfp <- df_long %>%
filter(RL_period == RLp) %>%
mutate(RL_value = if (which_scale == "neg") RL_neg else RL_speed) %>%
filter(is.finite(RL_value))

if (nrow(dfp) == 0) stop("No finite values for RL", RLp, " on scale ", which_scale)

y_lo <- quantile(dfp$RL_value, 1 - zoom_prob, na.rm = TRUE)
y_hi <- quantile(dfp$RL_value, zoom_prob,     na.rm = TRUE)

scale_label <- if (which_scale == "neg") "-time" else "speed"

ggplot(dfp, aes(x = factor(years), y = RL_value)) +
geom_boxplot(
width         = 0.6,
linewidth     = 0.4,
outlier.alpha = 0.10,
outlier.size  = 0.5
) +
coord_cartesian(ylim = c(y_lo, y_hi)) +
scale_x_discrete(breaks = as.character(x_breaks)) +
scale_y_continuous(labels = scales::label_number(accuracy = 0.005)) +
labs(
title = paste0("Convergence of RL", RLp, " (", scale_label, " scale)"),
x     = "Record length (years)",
y     = paste0("Estimated return level (", scale_label, ")")
) +
theme_minimal(base_size = 14) +
theme(
panel.grid.minor   = element_blank(),
panel.grid.major.x = element_blank(),
axis.text.x        = element_text(size = 11),
axis.text.y        = element_text(size = 11),
axis.title         = element_text(size = 12),
plot.title         = element_text(face = "bold", size = 15),
plot.margin        = margin(10, 12, 10, 12)
)
}

plot_RL_box_zoom(results_long, RLp = 25, which_scale = "neg")
plot_RL_box_zoom(results_long, RLp = 25, which_scale = "speed")
```


“Truth” via long-run simulation
```{r}
years_true <- 100000

sim_true <- simulate_block_maxima_normal(
n_years    = years_true,
n_per_year = n_per_year,
mu_T       = mu_T,
sd_T       = sd_T
)

fit_true <- fit_gev_and_RL(sim_true$neg_time, periods = periods)

true_pars_neg <- c(mu = fit_true$mu, sigma = fit_true$sigma, xi = fit_true$xi)
true_RL_neg   <- fit_true$RL
true_RL_speed <- 40 / (-true_RL_neg)

cat("\nApproximate 'truth' parameters on -time scale (long-run fit):\n")
print(true_pars_neg)

truth_tbl <- tibble(
RL_period  = periods,
true_neg   = true_RL_neg,
true_speed = true_RL_speed
)

print(truth_tbl)
```
Bias/RMSE summaries
```{r}
bias_rmse_all <- results_long %>%
left_join(truth_tbl, by = "RL_period") %>%
mutate(
err_neg   = RL_neg   - true_neg,
err_speed = RL_speed - true_speed
) %>%
group_by(RL_period, years) %>%
summarise(
bias_neg   = mean(err_neg,   na.rm = TRUE),
rmse_neg   = sqrt(mean(err_neg^2,   na.rm = TRUE)),
sd_neg     = sd(RL_neg,      na.rm = TRUE),
bias_speed = mean(err_speed, na.rm = TRUE),
rmse_speed = sqrt(mean(err_speed^2, na.rm = TRUE)),
sd_speed   = sd(RL_speed,    na.rm = TRUE),
.groups    = "drop"
)

print(head(bias_rmse_all, 30))
```
Bias and RMSE plots
```{r}
plot_rmse_clean <- function(df_bias_rmse, RLp, which_scale = c("neg", "speed"),
n_min_rate = 25,
x_breaks = c(5, 10, 20, 30, 40, 50, 75, 100),
r2_min = 0.50) {
which_scale <- match.arg(which_scale)
dfp <- df_bias_rmse %>% dplyr::filter(RL_period == RLp)

rmse_col    <- if (which_scale == "neg") "rmse_neg" else "rmse_speed"
scale_label <- if (which_scale == "neg") "-time" else "speed"

dfp <- dfp %>% dplyr::mutate(rmse_plot = .data[[rmse_col]])

df_rate <- dfp %>%
dplyr::filter(years >= n_min_rate, is.finite(rmse_plot), rmse_plot > 0)

fit   <- lm(log(rmse_plot) ~ log(years), data = df_rate)
alpha <- -coef(fit)[["log(years)"]]
C     <- exp(coef(fit)[["(Intercept)"]])
r2    <- summary(fit)$r.squared

fit_label <- if (is.finite(alpha) && alpha > 0 && is.finite(r2) && r2 >= r2_min) {
paste0(
"Asymptotic fit (years >= ", n_min_rate, "): RMSE ≈ ",
sprintf("%.3g", C), " * n^{-", sprintf("%.2f", alpha), "}"
)
} else {
paste0(
"No clear power-law decay (years >= ", n_min_rate, "): ",
"alpha = ", sprintf("%.2f", alpha), ", R^2 = ", sprintf("%.2f", r2)
)
}

ggplot(dfp, aes(x = years, y = rmse_plot)) +
geom_line(linewidth = 0.6, alpha = 0.8) +
geom_point(size = 2, alpha = 0.8) +
scale_y_log10(
labels = scales::label_number(accuracy = 0.001),
breaks = scales::log_breaks(n = 6)
) +
scale_x_continuous(breaks = x_breaks) +
labs(
title = paste0("RMSE vs Record Length for RL", RLp, " (", scale_label, ")\n", fit_label),
x = "Record length (years)",
y = "RMSE (log scale)"
) +
theme_minimal(base_size = 14) +
theme(
panel.grid.minor = element_blank(),
plot.title = element_text(face = "bold")
)
}

plot_bias_clean <- function(df_bias_rmse, RLp, which_scale = c("neg", "speed"),
x_breaks = c(5, 10, 20, 30, 40, 50, 75, 100)) {
which_scale <- match.arg(which_scale)
dfp <- df_bias_rmse %>% dplyr::filter(RL_period == RLp)

bias_col    <- if (which_scale == "neg") "bias_neg" else "bias_speed"
scale_label <- if (which_scale == "neg") "-time" else "speed"

dfp <- dfp %>% dplyr::mutate(bias_plot = .data[[bias_col]])

ggplot(dfp, aes(x = years, y = bias_plot)) +
geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
geom_line(linewidth = 0.6, alpha = 0.8) +
geom_point(size = 2, alpha = 0.8) +
scale_x_continuous(breaks = x_breaks) +
scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
labs(
title = paste0("Bias vs Record Length for RL", RLp, " (", scale_label, ")"),
x = "Record length (years)",
y = "Bias"
) +
theme_minimal(base_size = 14) +
theme(
panel.grid.minor = element_blank(),
plot.title = element_text(face = "bold")
)
}

plot_rmse_clean(bias_rmse_all, RLp = 25, which_scale = "neg")
plot_rmse_clean(bias_rmse_all, RLp = 25, which_scale = "speed")
plot_bias_clean(bias_rmse_all, RLp = 25, which_scale = "neg")
plot_bias_clean(bias_rmse_all, RLp = 25, which_scale = "speed")
```

Convergence-rate and Delta-method table
```{r}
estimate_rate_and_constant <- function(df_bias_rmse,
which_scale = c("neg", "speed"),
n_min_rate = 25,
min_points = 4) {
which_scale <- match.arg(which_scale)
rmse_col <- if (which_scale == "neg") "rmse_neg" else "rmse_speed"

df2 <- df_bias_rmse %>%
mutate(rmse_tmp = .data[[rmse_col]]) %>%
filter(years >= n_min_rate, is.finite(rmse_tmp), rmse_tmp > 0)

df2 %>%
group_by(RL_period) %>%
group_modify(~{
dat <- .x
if (nrow(dat) < min_points) {
return(tibble(alpha = NA_real_, C = NA_real_, r2 = NA_real_, n_used = nrow(dat)))
}
fit <- lm(log(rmse_tmp) ~ log(years), data = dat)
tibble(
alpha  = -coef(fit)[["log(years)"]],
C      = exp(coef(fit)[["(Intercept)"]]),
r2     = summary(fit)$r.squared,
n_used = nrow(dat)
)
}) %>%
ungroup()
}

rate_neg   <- estimate_rate_and_constant(bias_rmse_all, which_scale = "neg",   n_min_rate = 25)
rate_speed <- estimate_rate_and_constant(bias_rmse_all, which_scale = "speed", n_min_rate = 25)

rate_summary <- rate_neg %>%
rename(alpha_neg = alpha, C_neg = C, r2_neg = r2, n_used_neg = n_used) %>%
left_join(
rate_speed %>%
rename(alpha_speed = alpha, C_speed = C, r2_speed = r2, n_used_speed = n_used),
by = "RL_period"
) %>%
mutate(
C_ratio     = C_speed / C_neg,
RL_neg_true = true_RL_neg[match(RL_period, periods)],
infl_theory = 40 / (RL_neg_true^2),

ratio_over_theory     = C_ratio / infl_theory,
alpha_theory          = 0.5,
alpha_neg_diff        = alpha_neg - alpha_theory,
alpha_speed_diff      = alpha_speed - alpha_theory,
alpha_neg_ratio       = alpha_neg / alpha_theory,
alpha_speed_ratio     = alpha_speed / alpha_theory,
alpha_speed_minus_neg = alpha_speed - alpha_neg,
alpha_speed_over_neg  = alpha_speed / alpha_neg

) %>%
arrange(RL_period)

print(rate_summary, n = nrow(rate_summary))
```
Write LaTeX table rows (robust in Knit AND interactive R)
```{r}
is_knitting <- isTRUE(getOption("knitr.in.progress"))

# Decide output directory
out_dir <- if (is_knitting) {
  # When knitting: write next to the Rmd so LaTeX \input finds it reliably
  tryCatch(
    normalizePath(dirname(knitr::current_input()), winslash = "/", mustWork = FALSE),
    error = function(e) normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  )
} else {
  # When running interactively: write to your project/data folder
  normalizePath(folder_path, winslash = "/", mustWork = FALSE)
}

# Fallback if folder_path doesn't exist (avoid accidental OneDrive surprises)
if (!dir.exists(out_dir)) {
  warning("Output directory does not exist: ", out_dir,
          "\nFalling back to current working directory: ", getwd())
  out_dir <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
}

out_tex <- file.path(out_dir, "rate_summary_rows_with_bottomrule.tex")

# Build the table
rate_tbl <- rate_summary %>%
  dplyr::transmute(
    p        = RL_period,
    alpha_mT = alpha_neg,
    C_mT     = C_neg,
    R2_mT    = r2_neg,
    alpha_V  = alpha_speed,
    C_V      = C_speed,
    R2_V     = r2_speed,
    ratio_C  = C_ratio,
    infl_th  = infl_theory
  ) %>%
  dplyr::arrange(p)

fmt <- function(x, d = 3) ifelse(is.na(x), "", formatC(x, format = "f", digits = d))

rate_tbl_fmt <- rate_tbl %>%
  dplyr::mutate(
    alpha_mT = fmt(alpha_mT, 3),
    C_mT     = fmt(C_mT, 3),
    R2_mT    = fmt(R2_mT, 3),
    alpha_V  = fmt(alpha_V, 3),
    C_V      = fmt(C_V, 3),
    R2_V     = fmt(R2_V, 3),
    ratio_C  = fmt(ratio_C, 3),
    infl_th  = fmt(infl_th, 3)
  )

lines <- apply(rate_tbl_fmt, 1, function(r) {
  paste0(
    r[["p"]], " & ",
    r[["alpha_mT"]], " & ", r[["C_mT"]], " & ", r[["R2_mT"]], " & ",
    r[["alpha_V"]],  " & ", r[["C_V"]],  " & ", r[["R2_V"]],  " & ",
    r[["ratio_C"]],  " & ", r[["infl_th"]], " \\\\"
  )
})

# Put \bottomrule INSIDE the file (since LaTeX tabular expects it there if you omit it outside)
lines <- c(lines, "\\bottomrule")

# Binary write (prevents BOM/encoding/control-char issues that can break booktabs)
con <- file(out_tex, open = "wb")
writeBin(charToRaw(paste(lines, collapse = "\n")), con)
close(con)

message("Wrote: ", normalizePath(out_tex, winslash = "/", mustWork = FALSE))
```
Reproducibility information
```{r}
sessionInfo()
```